---
title: "differential_expression"
output: html_document
date: '2024-03-01'
---

You may need to install some of these packages, you can try installing using 
conda or install them directly here. Conda and R packages sometimes do not
play nice with each other so I suggest you install them directly in R and then
record which versions using `sessionInfo()`. 

```{r setup, include=FALSE}
library(DESeq2)
library(tidyverse)
library(forcats)
library(data.table)
library(fgsea)
```


```{r}
counts <-  as.matrix(read.csv('results/verse_concat_filtered_vM34.csv', row.names = 'gene'))
head(counts)
```


I manually construct the coldata dataframe using `c()` and `rep()`. Ensure that
your samples are in the same order as they are found in your matrix. 

I make time a factor so that I can set the reference level. This is the
condition that will be the "baseline" that all fold changes are relative to. For
example, when we are comparing AD and P0 samples, a gene that has higher counts
in the AD samples will have a positive log2FoldChange.

This reference level is subjective and does not change the actual results but 
the interpretation of the results. It would be equally valid to set the reference
level as AD, but then you will have to remember that your log2foldchanges are
relative to the AD samples. 
```{r}
coldata <- data.frame(samples = colnames(counts), time = c(rep('P0', 2), rep('P4', 2), rep('P7', 2), rep('AD', 2)), row.names='samples')
coldata$time <- as.factor(coldata$time)
coldata$time <- relevel(coldata$time, ref='P0')
coldata
```

Here we call DESeq2 as described in the vignette. We use the `DESeqDataSetFromMatrix`
because we are starting with an external counts matrix.

When extracting the results, I specifically ask for the comparison between AD 
and P0 samples from the condition `time`. We set our timepoints as our condition
of interest in the coldata encoded in that same column, `time`. You could also
use this same argument, `results(contrast=c())` to pull out any other pairwise
comparison of interest between our timepoints.
```{r}
dds <- DESeqDataSetFromMatrix(countData = counts, colData = coldata, design = ~time)
dds <- DESeq(dds)
res <- results(dds, contrast=c('time', 'AD', 'P0'))
```


The DESeq2 vignette supplied this code and it simply arranges your results from
the above code block in order of padj descending. The second line simply converts
these results to a tibble for easier manipulation.
```{r}
resOrdered <- res[order(res$pvalue),]
resOrdered %>% as_tibble(rownames='geneid')
```

Here I just read in my delimited file containing the mapping of gene IDs to 
gene names as a tibble. 
```{r}
id2gene <- read_delim('results/id2gene.txt', col_names=c('geneids', 'genenames'))
```

I similarly convert the results to a tibble and give it a common column name
for the gene IDs. The same column name used in the id2gene tibble. This will
enable a `left_join()` which will preserve all of the original information from
`res` but also add in the associated gene names from id2gene. The `arrange()`
and `select` were just to make it more apparent that we now have the gene names
present in this same tibble. 
```{r}
res %>% 
  as_tibble(rownames='geneids') %>% 
  left_join(id2gene, by='geneids') %>% 
  arrange(padj) %>% 
  select(geneids, genenames, padj, log2FoldChange)
```
